---
title: "Reproducible Research: Peer Assessment 1"
author: "Steve Bryant"
date: "September 20, 2015"
output: html_document
    keep_md: yes
  pdf_document: default
---

## Loading and preprocessing the data

Download the file from the internet.

```{r get and load data}
if (!file.exists("activity.zip")){
      download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
                    destfile = "activity.zip", method = "wget")
      unzip("activity.zip")
}
activity <- read.csv("~/RepData_PeerAssessment1/activity.csv")
```


## What is mean total number of steps taken per day?

Calculate the total number of steps per day.

```{r # of steps/day}
steps <- aggregate(steps ~ date, sum, data=activity, na.action=na.omit)
```

Make a histogram of the total number of steps taken each day

```{r graph of steps}
hist(steps$steps, main="Average Steps Per Day", xlab="Steps Per Day")
hist(total.steps, breaks=10, main="Histogram of total steps per day", xlab="Steps Per Day", col="gray")
```

Calculate and report the mean and median total number of steps taken per day

```{r mean}
mean(steps$steps)
median(steps$steps)
```

## What is the average daily activity pattern?

```{r graph}
intervalData <- aggregate(steps ~ interval, mean, data=activity, 
                     na.action=na.omit)

plot(intervalData$interval, intervalData$steps, type="l", 
     main="Steps in 5 minutes", xlab="Intervals", ylab="Steps")
```

What interval did the subject usually take the most steps in?

```{r step/interval max}
intervalData[intervalData$steps == max(intervalData$steps),]
```

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r missing}
sum(is.na(activity$steps))
```

Create a new dataset that is equal to the original dataset but with the missing data filled in using average.

```{r fill in NA}
for(i in 1:length(activity$steps)) {
  if(is.na(activity[i,1])) {
    activity[i,1] <- intervalData[intervalData$interval == activity[i,3], 2]
  }
}
```

Re-graph the total number of steps per day with missing data filled in.
Does it make a big difference?
Is there any difference?

```{r regraph total steps}
steps <- aggregate(steps ~ date, sum, data=activity, na.action=na.omit)
hist(steps$steps, main="Average Steps Per Day", xlab="Steps Per Day")
```
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Mean of steps per day.

```{r mean2}
mean(steps$steps)
```

Median of steps per day.

```{r median2}
median(steps$steps)
```

Averages are about the same. There is a lower variance.

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r weekends weekdays}
activity$date <- as.Date(activity$date)

activity$day <- activity$steps

for(i in 1:length(activity$day)) {
  if(weekdays(activity[i,"date"]) == "Sunday") {
    activity[i,"day"] <- "weekend"
  }else if(weekdays(activity[i,"date"]) == "Saturday") {
    activity[i,"day"] <- "weekend"
  }else{
    activity[i,"day"] <- "weekday"
  }
}

activity$day <- as.factor(activity$day)

weekends <- activity[activity$day == "weekend",]
weekends <- aggregate(steps ~ interval, mean, data=weekends, 
                     na.action=na.omit)
weekends$day <- rep.int("weekend", length(weekends$steps))

weekdays <- activity[activity$day == "weekday",]
weekdays <- aggregate(steps ~ interval, mean, data=weekdays, 
                     na.action=na.omit)
weekdays$day <- rep.int("weekday", length(weekends$steps))

intervalData <- rbind(weekdays, weekends)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r plot time series}
library(lattice)
xyplot(steps ~ interval | day, data = intervalData, layout=c(1,2),
       main = "Steps per 5 minutes", xlab="interval", ylab="step",
       type="l")
```

Activity pattern is different between weekends and weekdays. Maybe they sit at a desk during the week.

Weekend average per 5 minute interval.
```{r weekend steps}
mean(weekends$steps)
```

Weekday average per 5 minute interval.
```{r weekday steps}
mean(weekdays$steps)
```